---
title: "Troubleshooting traits.build"
format: html
editor: visual
---

### Unsupported trait values

This error occurs when, for a categorical trait, the value in [data.csv]{style="color:blue;"} is different to the value in the traits dictionary ([config/traits.yml]{style="color:blue;"}).

```{r eval=TRUE}
table <- my_database$excluded_data %>%
  filter(dataset_id == current_study) %>%
  filter(error == "Unsupported trait value") %>%
  select(dataset_id, trait_name, value) %>%
  distinct()
```

You can individually add substitutions to [metadata.yml]{style="color:blue;"} using the function [metadata_add_substitution]{style="color:blue;"}

```{r eval=TRUE}
metadata_add_substitution(dataset_id = current_study, trait_name = "plant_growth_form", find = "T", replace = "tree")
```

Or, you can add an additional column to the table output (code above) and read it into [metadata.yml]{style="color:blue;"} using the function [metadata_add_substitutions_table]{style="color:blue;"}

The table read in must have the columns [dataset_id]{style="color:blue;"}, [trait_name]{style="color:blue;"}, [find]{style="color:blue;"}, and [replace]{style="color:blue;"}.

This is a hypothetical example for a table that contains 5 rows with plant_growth_form value that need updating.
```{r eval=TRUE}
table <- table %>%
  rename(find = value) %>%
  mutate(replace = c("tree", "mallee", "shrub", "graminoid", "herb"))

metadata_add_substitutions_table(table, dataset_id = dataset_id, trait_name = trait_name, find = find, replace = replace)
```

You can of course also write the table to a csv file, edit it in Excel or a text editor, then read it back into R.
```{r eval=TRUE}
write_csv(table, "data/dataset_id/raw/substitutions_required.csv")

...edit outside of R

table <- read_csv("data/dataset_id/raw/substitutions_required.csv")
```

### Dataset can't pivot wider

In order to convert a traits.build database into a wide format, the traits.build\$traits table must be able to pivot wider. This dataset was unable to pivot, due to duplication in the following rows:

```{r eval=TRUE}
my_database$traits %>%
  filter(dataset_id == dataset_ids) %>%
  select(dataset_id, trait_name, value, observation_id, source_id, taxon_name, entity_type, life_stage, value_type,
         population_id, individual_id, temporal_id, method_id, entity_context_id, original_name) %>%
  pivot_wider(names_from = trait_name, values_from = value, values_fn = length) %>%
  pivot_longer(cols = 14:ncol(.)) %>%
  rename(trait_name = name, number_of_duplicates = value) %>%
  select(dataset_id, taxon_name, trait_name, number_of_duplicates, observation_id, entity_type, value_type, population_id, everything()) %>%
  filter(number_of_duplicates > 1)
```

There are two likely explanations -- and solutions -- to this error:

**Note that individual-level observations are each assigned a unique `observation_id` and therefore should never create a problem with pivoting. Contact the AusTraits team if data with \`entity_type = "individual" are in the table of duplicates.**

1.  If your dataset combines individual (or population) level measurements with species-level measurements, the same species-level measurement may be read in many times. To solve this problem, you need to retain only the first instance of each species-level measurement, by including the following `custom_R_code`, where `taxon_name` is the column that contains taxon names and `column 1`, `column 2`, etc is a vector of the columns with categorical traits that require `de-duplicating`.

```{r eval=TRUE}
data %>%
  group_by(taxon_name) %>%
  mutate(across(c(`column 1`, `column 2`, `column 3`), replace_duplicates_with_NA))
  ungroup()
```

2.  Rows of data that represent measurements made at different times, 

    group_by(taxon_name)
